\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{textcomp}

\hypersetup{colorlinks=true, linkcolor=blue, urlcolor=blue, citecolor=blue}

\title{Methodology Appendix:\\International Football Elo Ratings\\and 2026 World Cup Predictions}
\author{Eric San Miguel}
\date{\today}

\begin{document}
\maketitle

\tableofcontents
\newpage

%----------------------------------------------------------------------
\section{Introduction}
%----------------------------------------------------------------------

This document provides a formal description of the methodology used to compute Elo ratings for men's and women's international football teams, as well as the Monte Carlo simulation framework used to generate 2026 World Cup predictions. The system is adapted from the World Football Elo Rating methodology published at \url{https://www.eloratings.net/about}, with modifications noted throughout.

%----------------------------------------------------------------------
\section{Elo Rating System}
%----------------------------------------------------------------------

\subsection{Overview}

Each team maintains a numerical rating that is updated after every international match. The rating change depends on (i) the importance of the match, (ii) the goal difference, (iii) the actual match result, and (iv) the expected result based on the pre-match rating difference between the two teams.

\subsection{Rating Update Formula}

After each match, a team's rating is updated according to:
\begin{equation}
R_{\text{new}} = R_{\text{old}} + K \cdot G \cdot (W - W_e)
\label{eq:elo_update}
\end{equation}
where:
\begin{itemize}[nosep]
    \item $R_{\text{new}}$ is the team's post-match rating,
    \item $R_{\text{old}}$ is the team's pre-match rating,
    \item $K$ is the tournament importance weight (Section~\ref{sec:kfactor}),
    \item $G$ is the goal difference multiplier (Section~\ref{sec:gfactor}),
    \item $W$ is the actual match result (Section~\ref{sec:matchresult}),
    \item $W_e$ is the expected match result (Section~\ref{sec:expected}).
\end{itemize}

The system is zero-sum: for any match between teams $A$ and $B$, the rating changes satisfy $\Delta R_A + \Delta R_B = 0$.

\subsection{Initial Rating}

All teams begin with an initial rating of $R_0 = 1500$. This rating is assigned the first time a team appears in the dataset. The Elo system is self-correcting; after approximately 20--30 matches, the initial rating has minimal influence on a team's current rating.

\subsection{Expected Result}
\label{sec:expected}

The expected result for team $A$ against team $B$ is given by:
\begin{equation}
W_e = \frac{1}{10^{-d_r / 400} + 1}
\label{eq:expected}
\end{equation}
where $d_r$ is the adjusted rating difference. For a match at a neutral venue:
\[
d_r = R_A - R_B
\]
For a match where team $A$ is the home team at a non-neutral venue:
\[
d_r = (R_A + H) - R_B
\]
where $H = 50$ is the home advantage parameter. The expected result for team $B$ is $1 - W_e$.

\paragraph{Home advantage.} A home advantage of $H = 50$ rating points is added to the home team's rating for the purpose of computing the expected result. This corresponds to approximately a 57\%--43\% advantage for the home team when teams are otherwise equally rated. Matches at neutral venues (e.g., World Cup group stage matches) receive no home advantage adjustment.

\paragraph{Remark.} The original eloratings.net system uses $H = 100$. We use $H = 50$ as we find this better calibrated to observed home advantage in international football.

\subsection{K Factor --- Tournament Importance}
\label{sec:kfactor}

The K factor determines the maximum possible rating change from a single match and varies by tournament type. Higher values are assigned to more prestigious competitions.

\begin{table}[h]
\centering
\begin{tabular}{@{}clp{7cm}@{}}
\toprule
$K$ & Category & Examples \\
\midrule
60 & World Cup \& Olympics & FIFA World Cup, Olympic Games \\
50 & Continental Championships & UEFA Euro, Copa Am\'erica, AFC Asian Cup, Gold Cup, Confederations Cup, Finalissima \\
40 & Qualifiers \& Nations Leagues & World Cup qualification, Euro qualification, UEFA Nations League, CONCACAF Nations League \\
30 & Other Tournaments & Regional championships, invitational cups, multi-sport games \\
20 & Friendlies & International friendly matches \\
\bottomrule
\end{tabular}
\caption{K factor assignment by tournament type.}
\label{tab:kfactor}
\end{table}

Tournament names are classified using the following precedence:
\begin{enumerate}[nosep]
    \item Exact match against a set of known World Cup/Olympic tournament names ($K=60$).
    \item Exact match against a set of known continental championship names ($K=50$).
    \item Exact match against a set of known qualifier tournament names, \textit{or} the tournament name contains the substring ``qualifi'' (case-insensitive) ($K=40$).
    \item Exact match against ``Friendly'' ($K=20$).
    \item Default: all other named tournaments ($K=30$).
\end{enumerate}

\subsection{G Factor --- Goal Difference Multiplier}
\label{sec:gfactor}

The goal difference index $G$ amplifies rating changes for decisive victories:
\begin{equation}
G = \begin{cases}
1.0 & \text{if } |g| \leq 1 \\
1.5 & \text{if } |g| = 2 \\
\dfrac{11 + |g|}{8} & \text{if } |g| \geq 3
\end{cases}
\label{eq:gfactor}
\end{equation}
where $g = \text{home\_score} - \text{away\_score}$ is the goal difference. Selected values are shown in Table~\ref{tab:gfactor}.

\begin{table}[h]
\centering
\begin{tabular}{@{}cc@{}}
\toprule
$|g|$ & $G$ \\
\midrule
0--1 & 1.000 \\
2 & 1.500 \\
3 & 1.750 \\
4 & 1.875 \\
5 & 2.000 \\
10 & 2.625 \\
\bottomrule
\end{tabular}
\caption{Goal difference multiplier values.}
\label{tab:gfactor}
\end{table}

\subsection{Match Result}
\label{sec:matchresult}

The actual result $W$ is coded as:
\[
W = \begin{cases}
1.0 & \text{win} \\
0.5 & \text{draw} \\
0.0 & \text{loss}
\end{cases}
\]

Matches decided by penalty shootout are treated as draws ($W = 0.5$ for both teams). Only the result in regulation and extra time determines the match outcome for rating purposes; the penalty shootout winner is disregarded.

%----------------------------------------------------------------------
\section{Data}
%----------------------------------------------------------------------

\subsection{Data Sources}

Match data is sourced from publicly available datasets maintained by Mart J\"urisoo, licensed under CC0 (public domain):
\begin{itemize}[nosep]
    \item \textbf{Women's:} \url{https://github.com/martj42/womens-international-results} --- 11,000+ matches from 1956 to present.
    \item \textbf{Men's:} \url{https://github.com/martj42/international_results} --- 49,000+ matches from 1872 to present.
\end{itemize}

Each dataset consists of a \texttt{results.csv} file containing match-level observations with the following fields: date, home team, away team, home score, away score, tournament name, city, country, and neutral venue indicator. A supplementary \texttt{shootouts.csv} file records penalty shootout outcomes.

\subsection{Team Name Normalization}

To maintain rating continuity across successor states, the following mappings are applied prior to computation:

\begin{table}[h]
\centering
\begin{tabular}{@{}ll@{}}
\toprule
Historical Name & Mapped To \\
\midrule
Macedonia & North Macedonia \\
FR Yugoslavia & Serbia \\
Serbia and Montenegro & Serbia \\
\bottomrule
\end{tabular}
\caption{Successor state name normalization.}
\end{table}

\subsection{Ranking Filters}

The published rankings exclude:
\begin{itemize}[nosep]
    \item Teams with fewer than 20 matches played.
    \item Teams whose most recent match occurred before January 1, 2016 (inactive/defunct teams).
    \item Non-FIFA member entities (64 excluded teams including CONIFA members, overseas territories, and historical entities).
\end{itemize}

%----------------------------------------------------------------------
\section{Derived Outputs}
%----------------------------------------------------------------------

\subsection{Smoothed Ratings (Trend)}

A smoothed rating series is computed for each team using a centered rolling average with a window of 365 days. The procedure is:
\begin{enumerate}[nosep]
    \item Construct a time series of post-match ratings indexed by match date.
    \item Resample to daily frequency using forward-fill interpolation.
    \item Apply a centered rolling mean with window size 365 and minimum periods 1.
    \item Extract smoothed values at the original match dates.
\end{enumerate}

\subsection{Ranking History}

Team rankings are computed at each match date from 1990 onward (1900 for men's football). At each date, all teams with recorded ratings are sorted in descending order by rating, and ranks are assigned. Historical ranking snapshots are also computed at monthly intervals (1st of each month) for the top 50 teams.

%----------------------------------------------------------------------
\section{2026 World Cup Predictions}
\label{sec:worldcup}
%----------------------------------------------------------------------

\subsection{Overview}

Tournament predictions are generated via Monte Carlo simulation of the complete 2026 World Cup. Each simulation independently samples the entire tournament---group stage, third-place qualification, and all knockout rounds---and the final probabilities are computed as the fraction of simulations in which each team reached each stage. The simulation uses $N = 10{,}000$ independent iterations.

\subsection{Score Prediction Model}

Match scores are predicted using an Elo-calibrated Poisson model. For each team, the expected number of goals is:
\begin{equation}
\lambda_A = \mu \cdot \exp\!\big(c \cdot d_r + c_2 \cdot d_r^2\big), \qquad \lambda_B = \mu \cdot \exp\!\big(-c \cdot d_r + c_2 \cdot d_r^2\big)
\label{eq:poisson}
\end{equation}
where $d_r = (R_A + H_A) - R_B$ is the adjusted rating difference (including home advantage $H = 50$ for the home team, $0$ for neutral venues). The deployed parameters are $\mu = 1.241$ (baseline expected goals per team), $c = 2.17 \times 10^{-3}$ (linear Elo scaling), and $c_2 = -5.25 \times 10^{-7}$ (quadratic term). The negative $c_2$ encodes sublinearity at extreme rating gaps---without it, the linear model extrapolates poorly into the high-$|d_r|$ regime that the rating-uncertainty Monte Carlo (Section~\ref{sec:wcuncertainty}) can produce.

\paragraph{Calibration.} Parameters were estimated by Poisson GLM with log link, fit on 64,202 team-match rows constructed by replaying Elo on men's international results from 1990 onward (49,215 matches; the dataset includes friendlies). Each match contributes two rows---one per team perspective---with $d_r$ taking opposite signs. Standard errors are clustered by match. Adding $d_r^2$ to a linear-only specification reduces AIC by 552 and substantially improves the calibration of the upper tail. A separate home-team dummy is also significant in the goal model ($+0.122$ log-goals, $z = 15$), but production code keeps the symmetric two-coefficient form $\lambda(d_r)$ because home advantage is already encoded at the rating layer through $H = 50$. The full fit specification, including the diagnostic with the home dummy, is implemented in \texttt{src/football\_elo/calibrate\_poisson.py}.

Goals for each team are sampled independently from Poisson distributions:
\[
\text{goals}_A \sim \text{Poisson}(\lambda_A), \qquad \text{goals}_B \sim \text{Poisson}(\lambda_B)
\]

Win, draw, and loss probabilities are derived analytically by summing over all possible scorelines:
\begin{align}
P(\text{win}_A) &= \sum_{i > j} P(\text{Pois}(\lambda_A) = i) \cdot P(\text{Pois}(\lambda_B) = j) \\
P(\text{draw}) &= \sum_{i} P(\text{Pois}(\lambda_A) = i) \cdot P(\text{Pois}(\lambda_B) = i)
\end{align}

For evenly matched teams at a neutral venue ($d_r = 0$), the model produces $P(\text{draw}) \approx 27\%$, consistent with the empirical draw rate in international football.

\subsection{Group Stage Simulation}

For each of the 12 groups, the six pairwise matches are simulated independently using the Poisson score model. Points are awarded as 3 for a win, 1 for a draw, and 0 for a loss. Goal difference and goals scored are computed directly from the sampled scorelines.

Teams within each group are ranked by:
\begin{enumerate}[nosep]
    \item Points (descending)
    \item Goal difference (descending)
    \item Goals scored (descending)
    \item Random tiebreaker (uniform)
\end{enumerate}

\subsection{Third-Place Qualification}

From each group, the third-place team is recorded with its points total and goal difference. The 12 third-place teams are ranked across all groups by:
\begin{enumerate}[nosep]
    \item Points (descending)
    \item Goal difference (descending)
    \item Random tiebreaker (uniform)
\end{enumerate}
The top 8 third-place teams advance to the Round of 32.

\subsection{Round of 32 Bracket}

The Round of 32 consists of 16 matches with predetermined pairings based on group finishing positions. Eight matches involve group winners against third-place qualifiers; four matches pair runners-up from different groups; and four matches pair group winners against runners-up from other groups.

The specific bracket assignment is shown in Table~\ref{tab:r32}.

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}cll@{}}
\toprule
Match & Team A & Team B \\
\midrule
73 & Runner-up A & Runner-up B \\
74 & Winner E & 3rd-place slot 0 \\
75 & Winner F & Runner-up C \\
76 & Winner C & Runner-up F \\
77 & Winner I & 3rd-place slot 1 \\
78 & Runner-up E & Runner-up I \\
79 & Winner A & 3rd-place slot 2 \\
80 & Winner L & 3rd-place slot 3 \\
81 & Winner D & 3rd-place slot 4 \\
82 & Winner G & 3rd-place slot 5 \\
83 & Runner-up K & Runner-up L \\
84 & Winner H & Runner-up J \\
85 & Winner B & 3rd-place slot 6 \\
86 & Winner J & Runner-up H \\
87 & Winner K & 3rd-place slot 7 \\
88 & Runner-up D & Runner-up G \\
\bottomrule
\end{tabular}
\caption{Round of 32 bracket structure. Matches are paired consecutively for the Round of 16 (i.e., winners of matches 73 and 74 meet in R16, etc.).}
\label{tab:r32}
\end{table}

\paragraph{Third-place slot allocation.} The eight qualifying third-place teams are assigned to bracket slots subject to the constraint that a third-place team cannot face the group winner from its own group. The allocation proceeds greedily in slot order: for each slot, the first available third-place team (sorted by group letter) that does not violate the same-group constraint is assigned. If no valid assignment exists, the constraint is relaxed.

\paragraph{Remark.} The official FIFA regulations specify 495 possible allocation scenarios based on which eight groups produce qualifying third-place teams. Our simplified greedy allocation approximates this system. The resulting tournament-level probabilities (e.g., $P(\text{winner})$) are robust to the specific allocation rule, as the primary determinant of advancement is team strength rather than bracket position.

\subsection{Knockout Stage Simulation}

Knockout matches use the same Poisson score model (Eq.~\ref{eq:poisson}). If the sampled scores produce a draw, the match is decided by a penalty shootout, where the winning probability is the Elo expected result:
\begin{equation}
P(\text{shootout win}_A) = W_e = \frac{1}{10^{-(R_A + h - R_B)/400} + 1}
\end{equation}
where $h = H = 50$ if team $A$ is a host nation, $h = -H$ if team $B$ is a host nation, and $h = 0$ otherwise.

The bracket progresses as a standard single-elimination tournament:
\[
\text{R32 (16 matches)} \to \text{R16 (8)} \to \text{QF (4)} \to \text{SF (2)} \to \text{Final (1)}
\]

Winners of consecutive R32 match pairs meet in the R16, and so forth.

\subsection{Output Probabilities}

For each of the 48 teams, the simulation reports:
\begin{itemize}[nosep]
    \item $P(\text{1st})$, $P(\text{2nd})$, $P(\text{3rd})$, $P(\text{4th})$: group stage finishing position probabilities.
    \item $P(\text{R32})$: probability of advancing to the Round of 32.
    \item $P(\text{R16})$, $P(\text{QF})$, $P(\text{SF})$, $P(\text{Final})$: probability of reaching each knockout stage.
    \item $P(\text{Winner})$: probability of winning the tournament.
\end{itemize}

By construction, $\sum_{i=1}^{48} P_i(\text{Winner}) = 1$ and for each team, $P(\text{R32}) \geq P(\text{R16}) \geq P(\text{QF}) \geq P(\text{SF}) \geq P(\text{Final}) \geq P(\text{Winner})$.

%----------------------------------------------------------------------
\section{Composite Rating with Squad Strength and Rating Uncertainty}
\label{sec:squadstrength}
%----------------------------------------------------------------------

\subsection{Motivation}

National teams play only 8--12 matches per year, so Elo ratings can lag meaningful roster changes. A team whose squad has dramatically improved may carry an outdated rating, and aging superstars whose on-field contribution remains high are not distinguished from declining players by Elo alone. Additionally, a point-estimate Elo treats a team's current rating as its true strength, which overstates confidence: strong favorites win six straight matches in the Monte Carlo far more often than they do in reality. We address both issues by (i) blending Elo with a Transfermarkt-based squad strength score and (ii) injecting Gaussian rating uncertainty into each tournament simulation.

\paragraph{Status.} Both adjustments are applied to the 2026 World Cup predictions published on the live site, calibrated on the 2018 and 2022 men's World Cups. Deployed parameters: $\beta = 0.25$ (squad weight), $\sigma = 120$ Elo points (rating uncertainty).

\subsection{Data Sources}

Three datasets are joined:
\begin{itemize}[nosep]
    \item \textbf{World Cup rosters:} jfjelstul/worldcup \url{https://github.com/jfjelstul/worldcup} --- 13{,}843 player-tournament records covering all men's WCs 1930--2022.
    \item \textbf{Historical player market values:} dcaribou/transfermarkt-datasets \url{https://github.com/dcaribou/transfermarkt-datasets} --- 616{,}000 player-valuation records with dates.
    \item \textbf{Player metadata:} Same repository. Provides date of birth, nationality, and peak career TM value per player.
\end{itemize}

Players are matched across datasets via a three-stage procedure: (i) exact normalized full-name plus country match; (ii) last-name plus country, restricted to uniquely identifiable names within a country; (iii) token-subset match within country. The procedure achieves an 88\% overall match rate with 100\% for top European nations; match rates for Brazil, Iran, and Saudi Arabia are 60--75\% due to nickname and transliteration conventions.

\subsection{Age Adjustment}

TM market values discount aging players more aggressively than on-field performance actually declines --- resale value and contract length dominate the pricing of older players. To recover a performance-oriented proxy, we apply a two-step correction.

\paragraph{Step 1: Recover peak-equivalent talent.} Divide the current TM value by TM's age-discount factor:
\begin{equation}
L(v, a) = \frac{v}{d_{\text{TM}}(a)}
\label{eq:age_step1}
\end{equation}
where $d_{\text{TM}}(a)$ is the empirical mean ratio $\mathbb{E}[v(a) / v_{\text{peak}}]$ for a sample of players with long TM histories. The built-in approximation used in this implementation is:
\[
d_{\text{TM}}(a) = \begin{cases}
1.0 & a \leq 26 \\
1.0 - 0.06\,(a - 26) & 27 \leq a \leq 32 \\
\max\{0.10,\; 0.64 - 0.15\,(a - 32)\} & a \geq 33
\end{cases}
\]

\paragraph{Step 2: Apply on-field performance decay.} Multiply recovered talent by a shallower performance-age factor consistent with published soccer-aging literature:
\begin{equation}
\text{adj}(v, a) = L(v, a) \cdot p(a)
\label{eq:age_step2}
\end{equation}
with
\[
p(a) = \begin{cases}
\max\{0.85,\; 1.0 - 0.02\,(27 - a)\} & a \leq 27 \\
1.0 - 0.02\,(a - 27) & 28 \leq a \leq 32 \\
0.90 - 0.04\,(a - 32) & 33 \leq a \leq 36 \\
\max\{0.40,\; 0.74 - 0.07\,(a - 36)\} & a \geq 37
\end{cases}
\]

\paragraph{Example.} A 32-year-old valued at \texteuro 60M with a TM peak of \texteuro 100M has $d_{\text{TM}}(32) = 0.64$, giving $L = \texteuro 94$M. Applying $p(32) = 0.90$ yields $\text{adj} = \texteuro 85$M, close to his peak --- capturing that his on-field level remains near-peak despite TM's discount.

\subsection{Squad Score}

Raw TM values are highly right-skewed and exhibit diminishing marginal returns at the player level --- a \texteuro 100M player is not ten times better than a \texteuro 10M player, and the gap between an amateur-level (\texteuro 0) and a serious professional (\texteuro 5--10M) is much larger than between a professional and a superstar. We therefore apply a $\log(1+x)$ transform per player before aggregating, with values measured in millions:
\begin{equation}
\tilde{s}(v, a) = \log\!\Bigl(1 + \frac{\text{adj}(v, a)}{10^6}\Bigr)
\label{eq:log_transform}
\end{equation}
This compresses the 110$\times$ raw team-mean spread observed in 2018 to roughly 10$\times$ in log space --- a more reasonable reflection of actual talent differences --- and naturally emphasizes the amateur-to-professional threshold over the professional-to-star jump.

A team's squad score is the \emph{mean} of log-transformed player scores across matched players:
\begin{equation}
S_T = \frac{1}{|\mathcal{P}_T|} \sum_{p \in \mathcal{P}_T} \tilde{s}(v_p, a_p)
\label{eq:squad_score}
\end{equation}
where $\mathcal{P}_T$ is the set of players in team $T$'s squad with valid TM data. Mean aggregation is robust to variable squad sizes across tournaments (23 in 2018, 26 in 2022) and to variable match rates across nations.

Squad scores are z-normalized across the teams in a given tournament:
\begin{equation}
z_T = \frac{S_T - \bar{S}}{\sigma_S}
\label{eq:squad_z}
\end{equation}

\subsection{Composite Rating}
\label{sec:squadcalibration}

The composite rating is a one-parameter blend of Elo and the z-normalized squad score:
\begin{equation}
R_T^{\text{composite}} = R_T^{\text{Elo}} + \beta \cdot z_T \cdot \sigma_{\text{Elo}}
\label{eq:composite}
\end{equation}
where $\sigma_{\text{Elo}}$ is the standard deviation of Elo across the tournament's participating teams. This rescaling keeps the squad bump in Elo-equivalent units. At $\beta = 0$ the model reduces to pure Elo; at $\beta = 1$ a one-$\sigma$ squad advantage contributes a full Elo-$\sigma$ boost.

\subsection{Rating Uncertainty}
\label{sec:ratingsigma}

Monte Carlo simulations that treat Elo ratings as point estimates systematically overstate the probability of top favorites: a team rated 200 Elo points above its nearest rival will be predicted to win a 90-minute knockout match roughly 65\% of the time, and compounding this across six rounds gives an unrealistic championship probability. In practice, a team's rating has genuine uncertainty --- form varies tournament-to-tournament, injuries reshape squads, and a single-match outcome is driven by factors Elo cannot capture.

We inject rating uncertainty by sampling each team's rating once per tournament simulation from a Gaussian centered on the composite rating:
\begin{equation}
R_T^{(s)} = R_T^{\text{composite}} + \varepsilon_T^{(s)}, \qquad \varepsilon_T^{(s)} \sim \mathcal{N}(0, \sigma^2)
\label{eq:rating_sigma}
\end{equation}
The sample $(s)$ is drawn at the start of each of the 10{,}000 tournament iterations and held fixed for all six matches in that iteration, so uncertainty is coherent within a simulated tournament. This structurally compresses top-of-tier probabilities: a fraction of simulations have Spain at 2100 rather than 2220, and in those simulations Spain does not steamroll the bracket.

Match-level probabilities displayed on the Groups tab are computed analogously --- each shown W/D/L is the average of \texttt{match\_probabilities} evaluated under 400 independent rating samples --- so the per-match bars on the site are self-consistent with the tournament-level aggregates.

\subsection{Calibration}

We grid-search $\beta$ and $\sigma$ jointly on the pooled 2018+2022 men's WC match outcomes, minimizing the mean multiclass Brier score:
\[
\mathcal{B}(p, y) = \sum_{k \in \{W, D, L\}} \bigl(p_k - \mathbb{1}\{y = k\}\bigr)^2
\]
applied to every match in the two tournaments.

\subsection{Results}

Pure-Elo match-level Brier baselines are shown in Table~\ref{tab:brier_baseline}. Both are well below the 0.667 uniform-prediction baseline, confirming Elo has predictive signal.

\begin{table}[h]
\centering
\begin{tabular}{@{}lcc@{}}
\toprule
Tournament & Brier & Log-loss \\
\midrule
2018 (Russia) & 0.5765 & 0.9739 \\
2022 (Qatar)  & 0.6130 & 1.0540 \\
\midrule
Uniform baseline & 0.667 & 1.099 \\
\bottomrule
\end{tabular}
\caption{Pure-Elo match-level scores on the two calibration World Cups.}
\label{tab:brier_baseline}
\end{table}

Univariate sweeps identify optima near $\beta \approx 0.2$ and $\sigma \approx 120$. A 2D grid search confirms a broad minimum around $\beta \in [0.2, 0.4], \sigma \in [100, 180]$, with the pooled Brier surface nearly flat across this region. The deployed parameters are $\beta = 0.25$ and $\sigma = 120$ --- the interior of the minimum region --- giving a pooled Brier of approximately 0.5864, or a 1.40\% relative improvement over pure Elo.

\paragraph{Diagnosis.} The log-transformed squad score $z_T$ has Pearson correlation of $0.79$ (2018) and $0.78$ (2022) with normalized Elo --- Elo already captures the bulk of squad-quality information. Rating uncertainty contributes orthogonal signal: it cannot be extracted from any rescaling of the Elo vector. The two adjustments therefore combine roughly additively rather than interacting strongly.

\paragraph{Tournament-level impact.} Match-level Brier is a conservative yardstick for the benefit of rating uncertainty because single-match predictions change only modestly; the large effect is on compounded tournament-level probabilities. With $\sigma = 0$, Spain's championship probability is 30.7\% and the top-3 teams combined hold 65\% of all tournament-winning mass. With $\sigma = 120$ and $\beta = 0.25$, Spain falls to 25.4\% and the top-3 combined share drops to 55.2\%, matching the 15--22\% top-favorite range and 40--55\% top-3 mass reported by consensus public models.

\paragraph{Statistical power.} With per-match Brier standard deviation $\approx 0.3$ and $n = 128$ matches, the standard error on the mean is $\approx 0.027$. A true 5\% lift ($\approx 0.030$ Brier points) would be detectable at roughly 50\% power; a 2\% lift ($\approx 0.012$ points) requires on the order of 1{,}100 matches. The backtest can rule out large effects but is underpowered for small ones. Adding 2006, 2010, and 2014 to calibration would triple the effective sample and sharpen the optimum.

\subsection{Deployed 2026 Output}

The deployed model applies $\beta = 0.25$ and $\sigma = 120$ to the current tournament. Current TM squads are constructed from players active in the 2023/24 season or later, ranked by market value and truncated at the top 26 per team. Table~\ref{tab:composite2026} contrasts pure Elo (what was shipping before) with the deployed composite + uncertainty model for the top contenders.

\begin{table}[h]
\centering
\begin{tabular}{@{}lrrrr@{}}
\toprule
Team & Elo & Composite & Pure $P(\text{Win})$ & Deployed $P(\text{Win})$ \\
\midrule
Spain     & 2218 & 2284 & 30.7\% & 25.4\% \\
Argentina & 2179 & 2230 & 19.6\% & 15.7\% \\
France    & 2137 & 2207 & 14.9\% & 14.1\% \\
England   & 2084 & 2154 & 6.4\%  & 8.3\%  \\
Brazil    & 2045 & 2108 & 3.5\%  & 5.1\%  \\
Portugal  & 2035 & 2091 & 2.9\%  & 4.2\%  \\
Germany   & 2009 & 2066 & 2.3\%  & 3.6\%  \\
Netherlands & 2028 & 2080 & 2.4\% & 3.4\% \\
Colombia  & 2046 & 2063 & 3.3\%  & 3.1\%  \\
Mexico    & 1960 & 1963 & 2.2\%  & 1.6\%  \\
\bottomrule
\end{tabular}
\caption{2026 WC win probabilities: pure Elo (pre-deployment) versus the deployed composite + uncertainty model ($\beta = 0.25$, $\sigma = 120$, log-transformed squad values). Top-3 combined share drops from 65.2\% to 55.2\%, in line with published consensus models.}
\label{tab:composite2026}
\end{table}

\subsection{Next Steps}

To improve statistical power and explore less-correlated signals, future iterations could:
\begin{itemize}[nosep]
    \item Extend calibration to 2006, 2010, and 2014 WCs (the data sources cover these years, roughly tripling the match sample).
    \item Extend to UEFA Euro 2016/2020/2024 and Copa Am\'erica editions for additional tournament-level observations.
    \item Fit the TM age-discount curve empirically via \texttt{fit\_tm\_age\_discount()} rather than using the built-in approximation.
    \item Replace or supplement TM values with starting-XI club Elo weighted by minutes --- a signal less directly captured by international Elo.
    \item Fit $\sigma$ per-team (more volatile teams get higher uncertainty) rather than a single tournament-wide value.
    \item Fit a logistic regression directly on match outcomes with EloZ and SquadZ as features, allowing the data to pick coefficients rather than a one-parameter blend.
\end{itemize}

%----------------------------------------------------------------------
\section*{References}
%----------------------------------------------------------------------

\begin{enumerate}[nosep]
    \item Elo, A. E. (1978). \textit{The Rating of Chessplayers, Past and Present.} Arco Publishing.
    \item World Football Elo Ratings. \url{https://www.eloratings.net/about}.
    \item J\"urisoo, M. Women's International Football Results. \url{https://github.com/martj42/womens-international-results}.
    \item J\"urisoo, M. International Football Results. \url{https://github.com/martj42/international_results}.
    \item Fjelstul, J. \textit{A Comprehensive Database on the FIFA World Cup.} \url{https://github.com/jfjelstul/worldcup}.
    \item dcaribou. \textit{Transfermarkt Datasets.} \url{https://github.com/dcaribou/transfermarkt-datasets}.
\end{enumerate}

\end{document}
